Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-07-16 paper1_jiturra.Rmd
appendix

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado  
#linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
abstract: | 
  | Rising economic inequality has increased attention to the link between class relations and redistributive preferences in contemporary societies. However, economic inequality unevenly affects class relations and their influence on support for redistribution across societies. While those in the upper and middle classes reinforce their privileged access to resources and diverse networks, the working class has been disproportionately affected by social exclusion as inequality rises. Consequently, inequality consolidates segregated lifeworlds as it increases the divide between the lives and experiences of the different social classes.  Meeting people from diverse class positions can help individuals become more aware of other people's lifestyles and worldviews. Homogeneous upper-class networks may reduce empathy and solidarity towards those in need, resulting in decreased support for redistribution. Conversely, segregated lower-class social networks can lead to increased marginalization and reinforce support for redistribution. This study explores the relationship between class-based network segregation and redistributive preferences employing cross-national data from 32,529 individuals in 31 societies. The main findings suggest that social class conditions the association of network homogeneity on redistributive preferences, where working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity decreases support for redistribution. In addition, the conditional influence of network segregation weakens in unequal societies, especially for the upper classes. Implications for the study of class relations and political attitudes are discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: yes
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa-no-doi-no-issue.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  # - \usepackage{times}           # Times New Roman
  - \usepackage{helvet}
  - \usepackage{endnotes}
  - \let\footnote=\endnote 
  - \renewcommand{\familydefault}{\sfdefault} 
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

```{r eval=FALSE, include=FALSE}
 # for render in pdf run rmarkdown::render_site("docs/paper.Rmd", output_format = "all")
 # clean #in the yml
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::pdf_document2")
 # rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::html_document2")
 rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::word_document2")

 # install.packages("trackdown")
 trackdown::update_file(file = "paper1_jiturra.Rmd", hide_code = TRUE,gpath = '001_PhD/002dissertation/paper1')

```

<!-- - considerar ISEI08 como alternativa de social class por la correlacion con homogeneidad -->
<!-- - darle una oportunidad al h-index -->
<!--     - en general es prácticamente lo mismo que el indicador de homogeneidad -->
<!-- - controlaria por social activities index () -->
<!-- - social resources  -->

@@

Over the past decades, cross-national studies on redistributive preferences and social class have predominantly focused on individuals and households [@lindh_class_2020]. In contrast, less attention has been paid to how social networks influence redistributive preferences. Sociologists have demonstrated that individuals tend to segregate social networks [@mcpherson_birds_2001;@lazarsfeld_friendship_1954]. Additionally, opinions about economic inequality can be affected by class segregation, as it provides insight into the economic conditions, lifestyles, and worldviews of other social classes [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. Despite this, the study of redistributive preferences can be enhanced by broadening the understanding of class relations from a network perspective. This approach offers a comprehensive understanding of processes such as shared norm internalization, class identity formation, and collective solidarity [@Svallfors2006;@goldthorpe_sobre_1992]. Thus, this study examines the relationship between class-based network segregation and redistributive preferences in a cross-national comparison, focusing on the role of economic inequality as a moderator in the influence of class relations on support for redistribution.

While prior research has predominantly examined the impact of social class through an individualistic lens, more attention should be devoted to understanding the role of social environments in class relations. This omission is particularly surprising given that class positions are fundamentally rooted in production relations, making them inherently relational, not only in their economic underpinnings but also in the power dynamics entwined within class conflicts [@wright_comparative_1989]. Furthermore, the normative basis of class relations introduces the relevance of solidarity and reciprocity dimensions, which are argued to provide the moral foundation for legitimacy and popular support for welfare schemes [@mau_moral_2003]. In this regard, beyond the influence of individual class position, recent efforts have highlighted that while the class positions of network ties do affect redistributive preferences, the degree of segregation in a homogeneous class environment can drive attitudes to become even stronger [@otero_lives_2022].

While network segregation widens the class divide in redistributive preferences, little is known about how income inequality can affect the association between class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link suggest that income inequality is crucial for understanding how class relations are reflected in redistributive demands, as it represents the current state of the distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding in this literature is that the upper classes tend to support redistribution more in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality leads to more segregated forms of social participation and network composition [@pichler_patterns_2007;@otero_space_2022]. Indeed, inequality reinforces stratified access to social activities and widens the distance between classes, resulting in increased marginalization of the lower classes and consolidating the privileged positions of the upper classes, who have better opportunities to diversify their social activities and interpersonal ties [@lancee_income_2012;@otero_differences_2023]. Nevertheless, current efforts to examine the impact of income inequality have primarily focused on either social networks or support for redistribution. A key question that remains unanswered is the extent to which broader economic disparities moderate the relationship between class-based network segregation and support for redistribution.

Thus, to the best of my knowledge, no research has comparatively investigated the role of economic inequality in the relationship between class-based network segregation and redistributive preferences. This study aims to address two key questions:

@@

For this investigation, I utilized a sample of 32,529 individuals across 31 countries from the International Social Survey Program (ISSP) 2017. This dataset provides unprecedented cross-national comparative data on social networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, social class not only represents individuals’ market situations but also their economic interests and moral perspectives on the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Here, redistributive preferences are understood as the support for policies and mechanisms aimed at reducing economic inequality through the redistribution of wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier to less wealthy individuals or groups in society [@garcia-sanchez_two_2022].

@@

Empirically, the current understanding of class divides in redistributive preferences is extensive [@lindh_public_2015;@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. Nevertheless, beyond the market situation of individuals and their households, considering class-based network ties highlights the relevance of the class situation, as it encompasses both economic interests and life chances that drive communal action beyond individuals [@weber_class_2011, pp. 57-59]. Thus, the role of social relations in shaping redistributive preferences from a network perspective considers more straightforwardly the social ties that structure class relations.

@@

<!-- It can be argued that class relations not only represent resource-based distinctions but also patterns of connectedness observed in the differentiation of social ties.  -->

Class relations can also be understood as the degree of cross-class relationships as a structural characteristic of society, representing a network of social ties between different class positions [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001, p.416]. In this context, friendship and family ties tend to be homogeneous concerning class and demographic characteristics, while more distant ties serve to bridge individuals to different social groups and contribute to network diversity [@diprete_segregation_2011; @lazarsfeld_friendship_1954]. Besides, socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004; @homans_human_1951]. Nevertheless, attitude similarity in segregated networks is rather a consequence of the social differentiation that structures the formation of cross-class network ties [@feld_focused_1981].

There are two primary approaches in the study of class relations from a network perspective. On the one hand, network diversity is defined as the degree of connectedness to dissimilar occupations, which represents vertical access to resources embedded in social networks [@lin_access_1986]. However, diversity is defined as the rate of dissimilar ties within the network without necessarily providing a reference position to describe network similarity. On the other hand, network segregation is the absence of cross-class connections and represents how similar network ties are in contrast to the individual. This perspective is conceptually closer to homophily as it is anchored in individual class positions and has been empirically addressed through network homogeneity [@otero_open_2021].

How does social class stratify social networks? Empirically, studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper-class forming connections with diverse individuals, in contrast to the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are observed in the composition of social ties, where the upper and intermediate classes maintain increasingly diverse and prestigious social environments compared to the working class [@cepic_how_2020;@carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest since class interests widen the social distance between proprietors and manual workers, the intermediate position of supervisors, along with their higher contact frequency with manual workers, makes the formation of friendship ties more likely. Similarly, @otero_open_2021 have shown that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class, suggesting that their limited life chances and lower capacity for social engagement ultimately result in a lack of social resources that leads to segregation. Conversely, the upper class is less permeable and homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

It has been stated that perceptions about economic inequality rooted in social comparison processes with reference groups can explain the formation of redistributive preferences [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. The argument posits that people form their beliefs through family, friends, and coworkers' experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of these reference groups [@Evans1992, p. 467]. Therefore, inferences about the social world are linked to the degree of segregation in the immediate social environment of a person, which influences the intensity and character of the information that ultimately shapes inequality perceptions [@mijs_is_2021]. Accordingly, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more segregated networks [@summers_deliberating_2022]. Nevertheless, I argue that this body of research has been mainly focused on the cognitive dimension of preference formation through inequality perceptions rather than straightforwardly addressing the claimed influence of network segregation on redistributive preferences [@cansunar_who_2021;@garcia-castro_perceived_2022].

In contrast, social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007, p. 550]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' class situations but also by the class positions of their network ties [@paskov_crossclass_2022]. Social influence processes can either align or divide opinions depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These findings reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences that shape broader social experiences [@morris_attenuation_1996, p. 48]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, spatially segregated interactions may lead individuals to view the lives of different classes as more distant, which can undermine feelings of social inclusion and cohesion [@sachweh_moral_2012]. Thus, segregation can exacerbate perceptions of others as strangers and may impact empathy and solidarity, potentially diminishing social cohesion [@otero_lives_2022, p. 758].

<!-- The attention has increased in recent years to how social ties are connected to attitudes toward inequality, and it has been shown that frequent interactions between classes can lead to different attitudinal changes depending on the individuals' social class. For instance, contact with the upper classes drives higher justification of economic inequality among the lower classes, whereas interactions with lower classes may prompt the different social classes to question the fairness of income distribution [@vargassalfate_contact_2023]. Further evidence suggests that class-based contact diversity shortens the social distance between social classes, providing broader images of the living conditions of others, which can trigger greater concerns about inequality [@otero_power_2023] and undermine support for the market distribution of social services [@beck_explorando_2019]. -->

The class position of surrounding family members, friends, and acquaintances influences redistributive preferences, as these individuals function as socialization agents whose impact can be amplified in segregated social networks. In principle, political attitudes are connected to class interests and norms that are nurtured in the family of origin during childhood and early adulthood. For instance, @lee_consider_2023 shows that individuals with network ties to the upper class through parental connections tend to support redistribution and progressive taxation less than those from working-class family backgrounds. Moreover, since households share risk based on the class position of their members, redistributive preferences are shaped not only by family background but also by the class positions of partners. For example, @paskov_crossclass_2022 indicates that working-class ties bolster redistributive preferences, whereas ties with the upper class decrease these preferences, with the effects becoming more pronounced when the class positions of individuals, partners, and parents create a more homogeneous network. Additionally, @lindh_missing_2021 found that friendship and acquaintanceship ties to the managerial class are associated with lower redistributive preferences compared to ties with the sociocultural and working classes. Hence, they suggest that individuals tend to adjust their attitudes based on the class position of their contacts [@lindh_missing_2021, p. 698].

@@

Studies on economic inequality, social class, and redistributive preferences indicate that the social classes react differently to rising economic inequality. Theoretically, political economists have suggested that high-income individuals are far from monolithic in their redistributive preferences, arguing that their concerns about the harmful _consequences_ of economic inequality (e.g., crime) ultimately motivate altruistic support for redistribution [@dimick_models_2018;@dimick_altruistic_2017;@rueda_externalities_2016]. Conversely, the moral economy literature in sociology has argued that the differences among the affluent can be explained as a matter of distributive justice evaluations about the _procedures_ for resource allocation [@liebig_sociology_2016].

Empirically, affluent groups are more responsive because they perceive the overall opportunity structure and social mobility chances are affected by rising inequality [@sachweh_why_2019]. Likewise, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of justice in the conditions for getting ahead [@kim_socioeconomic_2018]. In contrast, low-income individuals perceive ascribed characteristics as more important in constraining the opportunity structure, regardless of current income inequality [@sachweh_why_2019]. In a similar perspective, @edlund_democratic_2015 argue that in unequal societies with residual welfare states social class has a poor political meaning, making distributive conflicts weakly institutionalized around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can also be explained as they are often more aware of the institutional incentives and societal consequences of income inequality [@curtis_how_2015;@Svallfors2006].

How does income inequality affect social networks? As income inequality implies stronger barriers to establishing trustworthy social connections, this has also consequences on the structure of class relations. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] suggest that experiencing marginalization is deeply associated with diminished chances for participating openly in social life, which can be exacerbated in contexts of greater inequality. Also, societies with higher income inequality can erode social relations as they exhibit reduced social trust that operates as the basis for mutual understanding and feelings of solidarity [@kragten_income_2017]. Likewise, in contexts of lower income inequality and encompassing welfare states, the higher levels of civic and social participation increase generalized trust and reduce social conflict [@uslaner_inequality_2005]. Related evicende shows that higher community participation can incentivize support for redistribution in general, being particularly salient among high-income citizens [@yamamura_social_2012].

@@

The primary data source for this study is the "Social Networks and Social Resources" module of the International Social Survey Program (ISSP) [@issp_networks_2017]. The ISSP provides a nationally representative probability sample of the adult population in each participating country without substitution. Each country administers a carefully adapted questionnaire to ensure the cross-cultural validity of the data and enable meaningful comparisons between countries. The questionnaire includes sections on social networks, attitudes toward economic inequality, and demographic and socioeconomic background characteristics. The complete sample comprises 47,027 observations across 32 countries. However, after reviewing the required information and applying listwise case deletion, the final sample used in the analyses consists of 32,529 observations from 31 countries [^data].

@@

I use two indicators to measure redistributive preferences. The first indicator, support for government redistribution, is measured by the item: “It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes.” The second indicator, egalitarian preferences, is measured by the item: “For a society to be fair, differences in people’s standard of living should be small.” Both indicators use a five-point Likert scale with the following categories: ‘Strongly agree’ (1), ‘Agree’ (2), ‘Neither agree nor disagree’ (3), ‘Disagree’ (4), and ‘Strongly disagree’ (5), and they are shown to be correlated (r = 0.7). The indicators were reverse-coded, averaged, and normalized to a range from 0 to 100, where higher values reflect stronger redistributive preferences [@svallfors_government_2013].

@@

## Descriptive cross-country comparison on class, network segregation and redistributive preferences

>>>>

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.15'          # interlineado  
#linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
abstract: | 
  | Rising economic inequality has increased attention to the link between class relations and redistributive preferences in contemporary societies. However, economic inequality unevenly affects class relations and their influence on support for redistribution across societies. While those in the upper and middle classes reinforce their privileged access to resources and diverse networks, the working class has been disproportionately affected by social exclusion as inequality rises. Consequently, inequality consolidates segregated lifeworlds as it increases the divide between the lives and experiences of the different social classes. Meeting people from diverse class positions can help individuals become more aware of other people's lifestyles and worldviews. Homogeneous upper-class networks may reduce empathy and solidarity towards those in need, resulting in decreased support for redistribution. Conversely, segregated lower-class social networks can increase marginalization and reinforce redistribution support. This study explores the relationship between class-based network segregation and redistributive preferences, employing cross-national data from 31,694 individuals in 31 societies. The main findings suggest that social class conditions the association of network homogeneity on redistributive preferences, where working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity decreases support for redistribution. In addition, the conditional influence of network segregation weakens in unequal societies, especially for the upper classes. Implications for the study of class relations and political attitudes are discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: false
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

always_allow_html: true      
linkcolor: gray                         # enlaces y citas en color azul
bibliography: input/bib/rgroup-redist.bib     # bibliografia en bibtex
csl: input/bib/apa-no-doi-no-issue.csl
editor_options:
  chunk_output_type: console            # en RStudio, mostrar output en consola
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm" # márgenes de página
header-includes:
  - \usepackage{times}   # Times New Roman
  # - \usepackage{helvet}
  - \usepackage{endnotes}
  - \let\footnote=\endnote 
  # - \renewcommand{\familydefault}{\sfdefault}
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \setlength\parindent{24pt} 
  - \usepackage{fancyhdr}
  - \fancyhead{} 
---

```{r render, include=FALSE, eval=FALSE}
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::pdf_document2")
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::html_document2")
rmarkdown::render("paper1_jiturra.Rmd", output_format = "bookdown::word_document2")
```

@@

 the past decades, cross-national studies on redistributive preferences and social class have predominantly focused on individuals and households [@lindh_class_2020]. In contrast, less attention has been paid to the role of social networks on redistributive preferences. Sociologists have demonstrated that individuals tend to form segregated social networks according to social class [@mcpherson_birds_2001;@otero_lives_2022]. Additionally, attitudes toward economic inequality can be affected by class segregation, as it implies distance from other social classes'  economic conditions, lifestyles, and worldviews [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. Therefore, focusing on social ties can improve the understanding of the relationship between social class and redistributive preferences, as it broadens the scrutiny of class relations in terms of internalization of shared norms, class identity formation, and collective solidarity. [@Svallfors2006;@goldthorpe_service_1982]. In this discussion, the first part of this study examines the micro-level relationship between class-based network segregation and redistributive preferences. Subsequently, I focus on how cross-national differences in income inequality can moderate the relation of network segregation and support for redistribution.

While previous research has predominantly examined the impact of social class through an individualist lens, more attention has recently been paid to the role of social ties in class relations. This omission is especially surprising since social classes are rooted in social relations of production, which makes them inherently relational in their economic foundations and the power dynamics involved in distributive struggles [@wright_comparative_1989]. Moreover, the normative basis of class relations introduces the relevance of dimensions of solidarity and reciprocity, which are argued to provide the moral foundation for legitimacy and popular support for welfare schemes [@mau_moral_2003]. Thus, beyond the influence of individual class positions, recent efforts have highlighted that, while the class positions of network ties affect redistributive preferences, segregation in homogeneous class-based social settings can make attitudes even more pronounced [@otero_lives_2022].

As network segregation can widen the class divide in redistributive preferences, little is known about how income inequality may affect the association between class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link suggest that income inequality is crucial for understanding how class relations are reflected in redistributive demands, as it represents the current state of distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding in this literature is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality has implications for social participation and network composition. Indeed, inequality reinforces stratified access to social activities and widens the distance between classes, resulting in a higher diversity of networks and activities in the upper classes and more marginalization of the lower classes [@lancee_income_2012;@otero_differences_2023;@pichler_patterns_2007]. However, current efforts to examine the impact of income inequality have focused primarily on social networks or support for redistribution. Thus, a key question that remains unanswered is the extent to which higher income inequality moderates the relationship between social class-based network segregation and redistributive preferences.

To the best of my knowledge, no research has comparatively studied the role of economic inequality in the relationship between class-based network segregation and redistributive preferences. Therefore, this study aims to answer two key questions:

@@

In this study, I used a sample of 31,694 individuals from 31 countries from the International Social Survey Program (ISSP) 2017. This dataset provides unprecedented cross-national comparative data on social networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, social class not only represents individuals' labor market relations but also their economic interests and moral perspectives on the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Here, redistributive preferences are understood as the support for policies and mechanisms aimed at reducing economic inequality through the redistribution of wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier to less wealthy individuals or groups in society [@garcia-sanchez_two_2022].

@@

Empirically, the current understanding of class divides in redistributive preferences is extensive [@lindh_public_2015;@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. Nevertheless, considering class-based network ties highlights the relevance of the class situation, as it encompasses economic interests and life chances that drive communal action beyond individuals [@weber_class_2011, pp. 57-59]. Hence, considering the structure of social ties allows one to look more comprehensively into how class relations structure redistributive preferences.

@@

Class relations can also be comprehended as the structure of social ties between classes as a characteristic of the social structure, represented as networks formed by different social classes [@blau_macrosociological_1977]. Empirically, homophily in social relations is a consistent finding in the social network literature [@mcpherson_birds_2001]. This means that friendship and family ties tend to be homogeneous concerning social status and demographic characteristics, while distant ties bridge individuals to different social groups and contribute to network diversity [@diprete_segregation_2011;@lazarsfeld_friendship_1954]. Besides, socialization preferences indeed play a role in the formation of segregated networks [@visser_attitudes_2004;@homans_human_1951]. However, the similarity of attitudes in segregated networks results more from the social differentiation that structures class ties than from individual preferences [@feld_focused_1981].

There are two approaches in the study of class relations from a network perspective. On the one hand, network diversity is defined as the degree of connectedness to dissimilar occupations, which represents vertical access to resource diversity embedded in social networks [@lin_access_1986]. However, diversity is defined as the variability of different ties within the network without necessarily providing a reference position to describe network similarity. On the other hand, network segregation is the absence of cross-class connections and represents how similar network ties are in reference to the individual. This perspective is conceptually closer to homophily as it is anchored in individual class positions and has been empirically addressed through network homogeneity [@otero_open_2021].

How does social class stratify social networks? Empirically, studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper-class forming connections with diverse individuals, in contrast to the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are observed in the composition of social ties, where the upper and intermediate classes maintain increasingly diverse and prestigious social environments compared to the working class [@cepic_how_2020;@carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest that since class interests widen the social distance between proprietors and manual workers, the intermediate position of supervisors and their higher contact frequency with manual workers makes the formation of friendship ties more likely. Similarly, @otero_open_2021 found that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class, suggesting that their limited life chances and lower capacity for social engagement ultimately result in segregation. Conversely, the upper class is more homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

@@

The first approach states that perceptions about economic inequality rooted in social comparison processes with reference groups could explain the formation of redistributive preferences [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@kelley_class_1995]. The argument posits that people form their beliefs through family, friends, and coworkers' experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of these reference groups [@Evans1992, p. 467]. Therefore, inferences about the social world are linked to the degree of segregation in the immediate social environment of a person, which influences the intensity and character of the information that ultimately shapes inequality perceptions [@mijs_is_2021]. Accordingly, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared with people in more segregated networks [@summers_deliberating_2022]. Nevertheless, I argue that this body of research has been mainly focused on the cognitive dimension of preference formation through inequality perceptions rather than straightforwardly addressing the claimed influence of network segregation on redistributive preferences [@cansunar_who_2021;@garcia-castro_perceived_2022].

The second approach suggests that social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' social class but also by the class positions of their network ties [@paskov_crossclass_2022]. Thus, social influence processes can either align or divide opinions depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These arguments reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences [@morris_attenuation_1996]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, segregated interactions may lead individuals to perceive the lives of different classes as more distant and undermine feelings of social integration [@sachweh_moral_2012]. Thus, segregation can potentially diminish social cohesion as it exacerbates perceptions of others as strangers which in turn reduces empathy and solidarity [@otero_lives_2022].

How do network ties affect redistributive preferences? The class position of surrounding family members, friends, and acquaintances affects support for redistribution, as these ties function as socialization agents whose impact can be amplified in segregated social networks. In principle, political attitudes are connected to class interests and norms as they are socialized in the family of origin during childhood and early adulthood. For instance, @lee_consider_2023 shows that individuals with network ties to the upper class through parental connections tend to support redistribution and progressive taxation less than those from working-class family backgrounds. Moreover, since households share risk based on the class position of their members, redistributive preferences are shaped not only by family background but also by the class positions of partners. For example, @paskov_crossclass_2022 indicates that working-class ties bolster redistributive preferences, whereas ties with the upper class decrease them, with the effects becoming more pronounced when the class positions of individuals, partners, and parents form a more homogeneous network. Similarly, @lindh_missing_2021 found that friendship and acquaintanceship ties to the managerial class are associated with lower redistributive preferences compared to ties with the sociocultural and working classes. Hence, this suggests that individuals tend to adjust their attitudes based on the class position of their contacts [@lindh_missing_2021].

@@

Studies on economic inequality, social class, and redistributive preferences indicate that social classes react differently to rising economic inequality. Theoretically, political economists have suggested that high-income individuals are far from monolithic in their redistributive preferences, arguing that their concerns about the harmful _consequences_ of economic inequality (e.g., crime) ultimately motivate altruistic support for redistribution [@dimick_models_2018;@dimick_altruistic_2017;@rueda_externalities_2016]. Conversely, the moral economy literature in sociology has argued that the differences among the affluent can be explained as a matter of distributive justice evaluations about the _procedures_ for resource allocation [@liebig_sociology_2016].

Empirically, it has been shown that affluent groups are more responsive as they perceive the overall opportunity structure and social mobility chances are affected by rising inequality [@sachweh_why_2019]. Likewise, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of justice in the conditions for getting ahead [@kim_socioeconomic_2018]. In contrast, low-income individuals perceive ascribed characteristics as more important in constraining the opportunity structure, regardless of current income inequality [@sachweh_why_2019]. In a similar perspective, @edlund_democratic_2015 argue that in unequal societies with residual welfare states, social class has a poor political meaning, making distributive conflicts weakly institutionalized around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can also be explained as they are often more aware of the incentives and societal consequences of income inequality [@curtis_how_2015;@Svallfors2006].

How does income inequality affect social networks? As income inequality implies stronger barriers to establishing trustworthy social connections, this also has consequences on the structure of class relations. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] suggest that experiencing marginalization is deeply associated with diminished chances for participating openly in social life, which can be exacerbated in contexts of greater inequality. Also, societies with higher income inequality can erode social relations as they exhibit reduced social trust that operates as the basis for mutual understanding and feelings of solidarity [@kragten_income_2017]. Likewise, in contexts of lower income inequality and encompassing welfare states, higher levels of civic and social participation increase generalized trust and reduce social conflict [@uslaner_inequality_2005]. In a similar vein, @yamamura_social_2012 showed that redistributive preferences rise as community-level participation in social activities increases. Moreover, high-income citizens in these communities are specifically more likely to support collective solidarity through government redistribution.

@@

The primary data source for this study is the "Social Networks and Social Resources" module of the International Social Survey Program (ISSP) [@issp_networks_2017]. The ISSP provides a nationally representative probability sample of the adult population in each participating country without substitution. Each country administers a carefully adapted questionnaire to ensure the cross-cultural validity of the data and enable meaningful comparisons between countries. The questionnaire includes sections on social networks, attitudes toward economic inequality, and demographic and socioeconomic background characteristics. The complete sample comprises 47,027 observations across 32 countries. However, after reviewing the required information and applying listwise case deletion, the final sample used in the analyses consists of 31,694 observations from 31 countries [^data].

@@

I use two indicators to measure redistributive preferences. The first indicator corresponds to support for government redistribution and it is measured by the item: "It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes." The second indicator is egalitarian preferences and it is measured by the item: "For a society to be fair, differences in people's standard of living should be small." Both indicators use a five-point Likert scale with the following categories: 'Strongly agree' (1), 'Agree' (2), 'Neither agree nor disagree'(3), 'Disagree' (4), and 'Strongly disagree'(5). Following @svallfors_government_2013, I reverse-coded, averaged and normalized the indicators in a 0 to 100 index, where higher values reflect stronger redistributive preferences.

@@

## Descriptive cross-country comparison on class, network segregation, and redistributive preferences
